---
title: "Homework 3, Group B"
author: "Abdalghani, Demirbilek, Morichetti, Zambon"
date: "Spring 2020"
output:
  html_document:
    toc: yes
  beamer_presentation:
    highlight: tango
  include: null
  ioslides_presentation:
    highlight: tango
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: yes
  slide_level: 2
  slidy_presentation:
    fig.height: 3
    fig.width: 4
    highlight: tango
header-includes:
- \usepackage{color}
- \definecolor{Purple}{HTML}{911146}
- \definecolor{Orange}{HTML}{CF4A30}
- \setbeamercolor{alerted text}{fg=Orange}
- \setbeamercolor{frametitle}{bg=Purple}
institute: University of Udine & University of Trieste
graphics: yes
fontsize: 10pt
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.625, dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/')
library(MASS)
```
```{r setup, include=FALSE}
library(knitr)
local({
  hook_plot = knit_hooks$get('plot')
  knit_hooks$set(plot = function(x, options) {
    paste0('\n\n----\n\n', hook_plot(x, options))
  })
})
```


## LEC
### Exercise 1

Compute the bootstrap-based confidence interval for the $score$ dataset using the studentized method.


*solution :*



```{r basic 1, echo=TRUE}
#loading data 
score <- read.table("student_score.txt", header = TRUE)

# function fo compute the parameter of interest --> eignratio 
psi_fun <- function(data) {eig <- eigen(cor(data))$values
                           return(max(eig) / sum(eig))}
psi_obs <- psi_fun(score)
#psi_obs

# estimate the bootstrap-based standard error, SE_boot:
# and estimate the Standard error from each bootstrap sample using jackknife: 
n <- nrow(score) 
B <- 10^4
s_vect <- rep(0,B)
SE_jack <- rep(0,B)
s_tmp <- rep(0, n)


for(i in 1:B) 
{
  ind <- sample(1:n, n, replace = TRUE)
  s_vect[i] <- psi_fun(score[ind,])
  for(j in 1:n) {s_tmp[j] <- psi_fun(score[ind,][-j,])}
  SE_jack[i] <- sqrt(((n - 1)/n) * sum((s_tmp - mean(s_tmp))^2))
  
}
```

```{r basic 2, echo=TRUE}

SE_boot <- sd(s_vect)
wald_ci <- psi_obs + c(-1, 1) * 1.96 * SE_boot 
print("Wald-type confidence interval :")
wald_ci

hist.scott(s_vect, main = "Wald-type CI")
abline(v = psi_obs, col = 2)
abline(v = wald_ci, col = "blue")


#percentile method : 
perc_ci <- quantile(s_vect, prob=c(0.025, 0.975))
attr(perc_ci, "names") <- NULL
perc_ci
hist.scott(s_vect, main = "Percentile method")
abline(v = psi_obs, col = 2)
abline(v = perc_ci, col = "blue")


#studentized Method
z_vect <- (s_vect - psi_obs)/SE_jack

#studentized bootstrap confidence interval :
stud_ci <- psi_obs - SE_boot * quantile(z_vect, prob=c(0.975, 0.025))
attr(stud_ci, "names") <- NULL 
print("Studentized bootstrap confidence interval :")
stud_ci
hist.scott(s_vect, main = "Studentized method")
abline(v = psi_obs, col = 2)
abline(v = stud_ci, col = "blue")

```

Comparing the Studentized bootstrap confidence interval and the Wald-type interval, and taking the point estimate as a reference, as given in the lecture the percentile confidence interval is wider on the left side and shorter on the right side. However, the studentized bootstrap confidence interval is wider on both sides (left and right).


### Exercise 2

Compute bootstrap-based confidence intervals for the $score$ dataset using the $boot$ package.


*solution :*

```{r basic 3, echo=TRUE}
library(boot)

psi_fun_boot = function(x, indices){
  data <- x[indices,]
  eig <- eigen(cor(data))$values
  return(max(eig) / sum(eig))
}

#bootstrap variances needed for studentized intervals
psi_fun_boot_var = function(x, indices){
  data <- x[indices,]
  b <- boot(score, psi_fun_boot, 100)
  eig <- eigen(cor(data))$values
  return(c(max(eig) / sum(eig), var(b$t)))
}

score <- read.table("student_score.txt", header = TRUE)
boot_results <- boot(score, psi_fun_boot, 1000)
boot.ci(boot.out = boot_results, , type = c("perc","basic"))

boot_var_results <- boot(score, psi_fun_boot_var, 1000)
boot.ci(boot.out = boot_var_results, type = "stud")

```



### Lab : Exercise 10

Let y=(1,0,0,1,0,0,0,0,0,1,0,0,1,0) collect the results of tossing n=14 times an unfair coin, where 1 denotes heads and 0 tails, and p=Prob(yi=1). 


1. Looking at the Stan code for the other models, write a short Stan Beta-Binomial model, where p has a Beta(a,b) prior with a=3, b=3.


Beta-Binomial.stan file is : 
```
data{
  int N;
  int y;
  real<lower=0> alpha;
  real<lower=0> beta;
}
parameters{
  real p;
}
model{
  target+=binomial_lpmf(y|N, p);
  target+=beta_lpdf(p|alpha,beta);
}

```

2.extract the posterior distribution with the function  extract().


```{r basic 10, echo=TRUE }
library(rstan)
y <- c(1,0,0,1,0,0,0,0,0,1,0,0,1,0)
n <- length(y)


alpha <- 3 
beta <- 3 


data <- list(N=n , y=sum(y), alpha = 3, beta = 3)
fit <- stan(file = "Beta-Binomial.stan", data = data, chains = 4, iter = 2000)

sim <- extract(fit)
```







